---
title: "eqviz"
author: "Min Park"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{eqviz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette brieftly walks through how to use the package `eqviz` to clean and visualize NOAA earthquake data with different charting options. The dataset can be found here: https://www.ngdc.noaa.gov/nndc/struts/form?t=101650&s=1&d=1

The package has been designed in three sections
* Cleaning the data
* Charting the earthquakes data into a timeline using geom_timeline and geom_timeline_label geom_functions
* Visualizing location information using interactive Leaflet.

```{r setup}
library(eqviz)
```

## Cleaning the data

The cleaning part is composed of four functions including helper functions.


### eq_load_data

This is where raw data is gathered. To make use of the file structure, `readr::read_delimit()` function is used internally. For easier processing, column names and types are defined in this function. As a result, `dataframe` object is to be returned.

### helper_neg_dates

This helper function is made to support `eq_clean_data` function. My modularizing pre-processing of BC dates, `eq_clean_data` can simply the process. In R, intrinsic functions of packages including `lubridate` cannot process BC dates, so workaround is needed. Given chracter-type BC dates,this function mirrors BC date from the AD0 origin date and obtains expected datetime object of BC date.

```{r bc_dates}
neg_dates = helper_neg_dates('-2015-01-01')
neg_dates
class(neg_dates)
```

### eq_location_clean

This function shortens the name of the `location` by removing the `country` part of the `data`. Given a column to work with, this returns a modified dataset column. The process is basically done by two phases. Firstly, the name of the country and surrounding `:` symbol is removed using regular expressions. Next, the remaining location data is titlized. 

```{r location_clean}
dataset = data.frame(location_name = c("Japan: Tokyo",
                                       "China: Beijing",
                                       "US: California: San Francisco"))
dataset$locationname = eq_location_clean(dataset$location_name)
dataset
```

### eq_clean_data

This function finalizes the housekeeeping process of dataset including conversion of data types, creating new columns, and limitng the dataset only to necessary attributes. Specifically, `date` column is created combining `year`, `month`, and `day` columns. `latitude` and `longitude` columns are converted to numeric types for further use in leaflet. Finally, `id, year, data, country, location_name, longitude, latitude, deaths, total_deaths, eq_primary` are columns remaining by the end of the process. The modified dataset is returned as a result. 

```{r data_clean}
dataset = data.frame(id = c(1,10,5,250,3),
                     location_name = c("Peru: Lima", 
                                       "Mexico: Mexico City",
                                       "CUBA: HABANA",
                                       "SRI LANKA: IslanD",
                                       "US: Maine"),
                     country = c("Peru",
                                 "Mexico",
                                 "CUBA",
                                 "SRI LANKA",
                                 "US"),
                     year = c(1950,1999,-320,1569,1850),
                     month= c(5,3,NA,11,5),
                     day = c(28,17,NA,4,2),
                     hour = c(12,6,NA,2,5),
                     latitude= c(-12.5,20.2,34.1,8.567,46.975),
                     longitude = c(-77,-103.5,-75.5,81.233,-66.659),
                     deaths = c(NA,1,2,8,NA),
                     total_deaths = c(NA,1,NA,9,NA),
                     eq_primary = c(7.5,3.4,2.9,7.9,5.5)
                     )
dataset = eq_clean_data(dataset)
dataset$location_name = eq_location_clean(dataset$location_name)
dataset
```


## Visualzing on a timeline (GeomTimeline and GeomTimelineLabel)

The part explains newly created geoms including GeomTimeline and GeomTimelineLabel which cooperatively display earthquakes data on a timeline with location lables.


### GeomTimeline

`GeomTimeline` combines pointsGrob and polylineGrob to draw magnitudes of earthquakes on a timeline. The magnitude can be represented by the size of circle and the casulaty is by the darkness of the color.

### geom_timeline

`geom_timeline` layer function is the interface where users will supply data and aesthetics. Excluding dataset, only `x` axis value (`date`) is needed. Optionally, aesthetics include `y` axis value (`country`), `size`, `color`, `xmin` and `xmax` among others. When `xmin` and `xmax` values are not provided, the geom internally calcalates the range inferred from the data. Otherwise, the plot is limited to that range.

```{r geom_timeline, fig.width = 7, fig.height=3}
ggplot2::ggplot(data=dataset,ggplot2::aes(x=date,colour=deaths,size=eq_primary)) +
  geom_timeline() +
  ggplot2::scale_x_date(name='DATE') +
  ggplot2::scale_size_continuous(name='Richter scale value') +
  theme_eq_custom()
```

### GeomTimelineLabel

`GeomTimeline` combines textGrob and polylineGrob to indicate the location info of specific time on the line. This is solely to supplement the function of `GeomTimeline`.

### geom_timeline_label

`geom_timeline_label` layer function is the interface where users will supply data and aesthetics. Excluding dataset, `x` axis value (`date`), `label` value (`location_name`), and `n_max` (maximum number of top earthquakes magnitudes) are needed. Optionally, aesthetics include `y` axis value (`country`), `size`, `color`, `xmin` and `xmax` among others. When `xmin` and `xmax` values are not provided, the geom internally calcalates the range inferred from the data. Otherwise, the plot is limited to that range. The same range of `xmin` and `xmax` should be provided to make the desired graphics.

```{r geom_timeline_label, fig.width = 7, fig.height=3}
ggplot2::ggplot(data=dataset,ggplot2::aes(x=date,colour=deaths,size=eq_primary)) +
  geom_timeline() +
  geom_timeline_label(ggplot2::aes(label=location_name,n_max=3)) +
  ggplot2::scale_x_date(name='DATE') +
  ggplot2::scale_size_continuous(name='Richter scale value') +
  theme_eq_custom()
```

## Mapping the data (leaflet)

To map the data on a map, `leaflet` library is used to create two following functions.

### eq_map

The function `eq_map` is capable of patching the location info on a map, with interactive popup infos. Depending on the choice of input, different infos are fetched to popup text.

```{r eq_map, fig.width = 7, fig.height=3}
dataset
eq_map(dataset, annot_col='date')
```

### eq_create_label

This helper function pre-processes popup text for leaflet popup text. Using html tags, three infos can be fetched including `Location`, `Magnitude`, and `Total deaths`. Actual columns used in the dataset are `location_name`, `eq_primary`, and `total_deaths`.

```{r eq_create_label, fig.width = 7, fig.height=3}
dataset$popup_text = eq_create_label(dataset)
dataset
eq_map(dataset, annot_col='popup_text')
```
